\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:introduction}
%---background---
The credibility of empirical research in econometrics is fundamentally tied to the robustness of its findings. However, recent developments have shown that statistical results are frequently driven by incredibly small subsets of data, making them highly sensitive to specific observations \parencite{broderick2020automatic}. While researchers typically rely on some classical diagnostic tools or robust estimators to maintain integrity, both approaches have limitations. Diagnostic tools often fail to capture complex interactions such as joint influence and masking \parencite{chatterjee1986influential, rousseeuw1990unmasking}. Conversely, robust estimators prioritize controlling worst-case sensitivity to stabilize the overall model, rather than identifying the specific and minimal subsets of data that actually drive a particular statistical conclusion. \parencite{hampel1974influence}.
This study addresses this gap by utilizing the exact Most Influential Sets (MIS) tool to redefine how we assess empirical robustness \parencite{kuschnig2021hidden}.

%---problem_statement---
Traditionally, leave-one-out measures such as Cook's distance and DFBETAs quantify influence by measuring the change in parameter vectors when a single observation is deleted. While useful as local sensitivity metrics, this single-observation design has important limitations \parencite{cook2000detection}. Similarly, leverage metrics based on the diagonal elements of the hat matrix ($h_{ii}$) measure how extreme an observation is in the predictor space. However, these tools have a breakdown point of 1/n. This means that even a single extreme outlier can make the diagnostic unreliable \parencite{rousseeuw2003robust}. These limitations are primarily driven by two intertwined phenomena: masking and swamping \parencite{hadi1993procedures}. Because classical methods rely on the sample mean and variance-covariance matrix, a cluster of outliers can severely distort these foundational moments. This distortion not only hides the true influential cluster from detection (the masking effect) but also shifts the regression plane so drastically that legitimate observations are falsely flagged as anomalies (the swamping effect). Consequently, classical diagnostics misdirect the researcher's attention with false positives, while real influential sets remain hidden \parencite{zhao2019multiple, rousseeuw1990unmasking}.

Instead of detecting sensitivity, high-breakdown robust estimators are designed to accommodate it globally. Methods such as Least Trimmed Squares (LTS) and its modern extensions employ ``hard trimming'' by identifying the exact subset of observations that minimizes the residual sum of squares, discarding severe outliers \parencite{rousseeuw2011robust, xie2022partial}. Similarly, the MM-estimator utilizes a multi-stage procedure to secure the regression line against bad leverage points before applying a redescending loss function to maximize statistical efficiency \parencite{yohai1987high, avella2018influence}. However, while these estimators successfully resist arbitrary data contamination, they introduce a ``robustness illusion.'' Their mathematical objective is to broadly control for worst-case sensitivity and stabilize the global model, rather than to explicitly identify the specific, minimal subsets driving a researcher's hypothesis. By systematically down-weighting or trimming data that deviates from the main trend, these procedures inherently obscure valuable inferential information regarding the underlying sources of structural sensitivity. As highlighted by \textcite{riani2014monitoring}, standard robust procedures often act as black boxes where the researcher ``loses information on the effect that each unit, outlier or not, has on the final proposed estimate.'' Furthermore, theoretical bounds dictate that these estimators cannot achieve simultaneously high breakdown point and high efficiency under the normal model \parencite{hossjer1995exact}, leaving researchers to rely on arbitrary hyperparameter choices rather than exact structural identification.

Beyond masking and the informational loss of robust trimming, both classical and robust approaches share a key weakness: they lack formal statistical inference. In standard empirical practice, researchers identify influential observations using arbitrary heuristic thresholds—such as flagging Cook's distance values greater than $4/n$ or leverage scores exceeding $2p/n$ \parencite{belsley2005regression}. However, because these local metrics lack a defined asymptotic null distribution, they cannot generate formal $p$-values or facilitate rigorous hypothesis testing. Consequently, a researcher is unable to mathematically distinguish whether a highly influential subset represents a genuine structural failure of the model, or merely natural sampling variation from the tails of a distribution. To move beyond arbitrary ``rules of thumb,'' modern sensitivity analysis requires a rigorous inferential foundation, leveraging frameworks such as Extreme Value Theory (EVT), to formally test the exact statistical significance of influential sets.

%---objectives---
Given the limitations of local diagnostics and the informational loss of robust estimators, there is a need for a methodology that identifies structural sensitivity while providing formal statistical inference. To address the gap, this study uses the exact Most Influential Sets (MIS) framework, integrating algorithmic subset identification with Extreme Value Theory (EVT). Specifically, this research is guided by the following core objectives:

\begin{enumerate}
	\item \textbf{Theoretical Formulation:} The study first establishes a rigorous methodological benchmark. It systematically contrasts the MIS framework of \parencite{kuschnig2021hidden} against classical influence diagnostics and robust estimators, mathematically mapping how MIS resolves their respective vulnerabilities to masking and inferential information loss.
	
	\item \textbf{Monte Carlo Simulations:} The accuracy and convergence of the MIS algorithm are validated through a structured simulation design. The Data Generating Process (DGP) combines a grid of complex covariate and error distributions with mixed targeted outlier types. These outliers are scaled using the Median Absolute Deviation (MAD) to induce precise masking and breakdown conditions.
	
	\item \textbf{Empirical Applications:} The framework is then applied to real-world data. First, it audits foundational datasets in econometrics and finance to determine whether classical findings are structurally robust or driven by hidden influential sets. Second, it is applied to standard machine learning benchmarks to detect structural vulnerabilities obscured by aggregate performance metrics.
	
	% --- Extension Work---
	% \item \textbf{Exploratory Extension:} 
	% ---------------------
\end{enumerate}

%---structure---
The remainder of this thesis is organised as follows. Chapter 2 reviews the literature, covering the development of classical diagnostic tools, robust estimators, and the need for a set-level inferential framework. Chapter 3 introduces the methodology, detailing the MIS algorithm and its hypothesis testing framework under Extreme Value Theory (EVT). Chapter 4 presents a Monte Carlo simulation study, evaluating the scalability and accuracy of the framework under complex conditions including severe masking and swamping. Chapter 5 applies the methodology to econometric datasets and machine learning benchmarks. Finally, Chapter 6 concludes the study and outlines directions for future research.

\end{document}